Event Tracking Across Languages - Cross-Lngual Document Similarity and Event Tracking \todo{Better title...}

### 2015-03-20 rupnik2015.tex
restructured

\end{abstract}
\section{Overview}
Structure
\begin{itemize}
\item Introduction
\item Cross lingual document similarity (model)
\item  Event Detection
\item  Combing the approaches
\item Experiments
\item Conclusions
\end{itemize}
The story:
\begin{itemize}
\item Description of the problem - we would liek to understand events ans sequences of events, but our observables (e.g. news stories) occur in a variety of languages
\item We present an approach to tackle this problem using a combination of techniques....
\item Ultimately we present a tool showing the feasibility of our approach
\end{itemize}

@@

To cluster the articles, rather than rely on machine translations, we adopt well-established monolingual text mining tools to the multilingual setting. This represents the continuation of previous work \cite{nips, iti, nips2, sikdd} where we explored representations of documents which were valid over multiple languages.  The representations could be interpreted as multi-lingual topics, which were then used as proxies to compute cross-lingual similarities between documents. To learn the representations, we use Wikipedia as a training corpus. SIgnificantly, we do not only consider the major or \emph{hub} languages such as English, German, French, etc. which have significant overlap in article coverage, but also smaller languages (in terms of number of Wikipedia articles) such as Slovenian and Hindi, which may have a negligible overlap in article coverage. We can then define a similarity between any two articles regardless of language,  which allows us to cluster the articles according to topic. The underlying assumption being that articles describing the same event are similar  and will therefore be put into the same cluster.

@@

{\oldText{Taken from XLite paper}
Document retrieval is a well-established problem in data mining. There have been a large number of different approaches put forward in the literature. In this paper we concentrate on a specific setting: multi-lingual corpora. As the availability of multi-lingual documents has exploded in the last few years, there is a pressing need for automatic cross-lingual processing tools.

The prime example is Wikipedia - in 2001 the majority of pages were written in English, while in 2012, the percentage of English articles has dropped to 14\%. In this context, we look at how to find similar documents across languages. In particular, we do not assume the availability of machine translators, but rather try to frame the problem such that we can use well-established machine learning tools designed for monolingual text-mining tasks. This work represents the continuation of previous work \cite{nips, iti, nips2, sikdd} where we explored representations of documents which were valid over multiple languages.  The representations could be interpreted as multi-lingual topics, which were then used as proxies to compute cross-lingual similarities between documents. We look at a specific aspect of this problem: the distribution of articles across languages in Wikipedia is not uniform. While the percentage of English articles has fallen, English is still the largest language and one of \emph{hub} languages which not only have an order of magnitude more articles than other languages, but also many comparable articles with most of the other languages.

When doing document retrieval, we encounter two quite different situations. If for example, we are looking for a German article comparable to an English article, there is a large alignment between the document corpora (given by the intersection in articles in the two languages) making the problem well-posed. If, however, we look for a relevant Slovenian article given a Hindi article, the intersection is small, making the problem much harder. Since almost all languages have a large intersection in articles with \emph{hub} languages, the question we ask in this paper is: can we exploit hub languages to perform better document retrieval between non-hub languages? A positive answer would vastly improve cross-lingual analysis between less represented languages. In the following section, we introduce our representation followed by our experiments, which also shed light on the structural properties of the multilingual Wikipedia corpus.
}

@@

@@

\subsection{Multilingual dataset}

@@

\subsection{Algorithms}
\label{sec:algorithms}

We will present three approaches: a simple approach based on $k$-means clustering, a standard approach based on singular value decomposition and a new method that is applicable in more general
settings. All approaches are based on linear maps and we do not focus on alternatives, such as machine translation and probabilistic models. Applicability of machine translation relies
on the availability of training resources (strongly aligned translation corpora), which are not available for many world languages, but there exist comparable corpora. Probabilistic
methods \todo{provide reasons for not using them: e.g. difficulty with parameter estimation}.

\subsubsection{$k$-means}

@@

\subsubsection{Cross-Lingual Latent Semantic Indexing}

@@

The original sum of squared correlation is formulated as an unconstrained problem:
\begin{equation*}
  \begin{aligned}
    & \underset{w_i \in \RR^{k}}{\text{maximize}}
    & & \sum_{i < j}^m  \rho(w_i^T Y_i, w_j^T Y_j)^2.
\end{aligned}
\end{equation*}
We solve a similar problem by restricting $i=1$ and omit optimizing over non-hub language pairs.
Let $D_{i,i} \in \RR^{k \times k}$ denote the empirical covariance of $\mathcal{Y}_i$ and $D_{i,j}$ denote the empirical cross-covariance computed based on $\mathcal{Y}_i$ and $\mathcal{Y}_j$. We solve the following constrained (unit variance constraints) optimization problem:
\begin{equation}\label{squaredCorHubOriginal}
  \begin{aligned}
    & \underset{w_i \in \RR^{k}}{\text{maximize}}
    & & \sum_{i = 2}^m  \left(w_1^T D_{1,i} w_i \right)^2
    & \text{subject to}
    & & w_i^T D_{i,i} w_i = 1, \quad\forall i = 1,\ldots, m.
\end{aligned}
\end{equation}
The constraints $w_i^T D_{i,i} w_i$ can be simplified by using the Cholesky decomposition $D_{i,i} = K_i^T \cdot K_i$ and substitution: $y_i := K_i w_i$. By inverting the $K_i$ matrices and defining  $G_i := K_i^{-T} D_{1,i} K_i^{-1}$, the problem can be reformulated:
\begin{equation}\label{squaredCorHub}
  \begin{aligned}
    & \underset{y_i \in \RR^{k}}{\text{maximize}}
    & & \sum_{i = 2}^m  \left(y_1^T G_{i} y_i \right)^2
    & \text{subject to}
    & & y_i^T y_i = 1, \quad\forall i = 1,\ldots, m.
\end{aligned}
\end{equation}
A necessary condition for optimality is that the derivatives of the Lagrangian vanish. The Lagrangian of (\ref{squaredCorHub}) is expressed as:
$$  L(y_1, \ldots, y_m, \lambda_1, \ldots, \lambda_m) = \sum_{i = 2}^m  \left(y_1^T G_{i} y_i \right)^2 + \sum_{i=1}^m \lambda_i \left(y_i^T y_i - 1\right).$$
Stationarity conditions give us:
\begin{equation}\label{dLdx1}
 \frac{\partial}{\partial x_1} L = 0 \Rightarrow \sum_{i = 2}^m  \left(y_1^T G_{i} y_i \right) G_i y_i + \lambda_1 y_1,
\end{equation}
\begin{equation}\label{dLdxi}
\frac{\partial}{\partial x_i} L = 0 \Rightarrow \left(y_1^T G_{i} y_i \right) G_{i}^T y_1 + \lambda_i y_i = 0,~i > 1.
\end{equation}
Multiplying the equations (\ref{dLdxi}) with $y_i^T$ and applying the constraints, we can eliminate $\lambda_i$ which gives us:
\begin{equation}\label{eqy1yi}
G_{i}^T y_1 = \left(y_1^T G_{i} y_i \right) y_i,~i > 1.
\end{equation}
Plugging this into (\ref{dLdx1}), we obtain an eigenvalue problem:
$$\left( \sum_{i = 2}^m G_i G_{i}^T \right) y_1 + \lambda_1 y_1 = 0.$$
The eigenvectors of $\left( \sum_{i = 2}^m G_i G_{i}^T \right)$ solve the problem for the first language. The solutions for $y_i$ are obtained from (\ref{eqy1yi}): $y_i := \frac{G_{i}^T y_1}{\| G_{i}^T y_1 \|}$.
Note that the solution (\ref{squaredCorHubOriginal}) to can be recovered by: $w_i := K_i^{-1} y_i$. The linear transformation of the $w$ variables are thus expressed as:
$$W_1 = $$
$$W_i = $$

@@

\subsection{Evaluation}

\subsubsection{Link prediction}

\subsubsection{Hub language}

@@

\subsection{Evaluation}

@@

\section{Related work}

\subsection{Cross-lingual Document Similarity}

\subsection{Multilingual Latent Factor Models}

Ask Wray

\subsection{Cross-lingual Event Linking}

>>>>

\end{abstract}
\section{Overview}
Structure
\begin{itemize}
\item Introduction
\item System overview
\item Cross lingual document similarity (algorithms)
\item Event linking
\item Experiments
\item Conclusions
\end{itemize}
The story:
\begin{itemize}
\item Description of the problem - we would like to understand global news as sequences of events, but our observables (e.g. news stories) occur in a variety of languages
\item We present an approach to tackle this problem using a combination of techniques....
\item Ultimately we present a tool showing the feasibility of our approach
\end{itemize}

@@

To cluster the articles, rather than rely on machine translations, we adopt well-established monolingual text mining tools to the multilingual setting. This represents the continuation of previous work \cite{nips, iti, nips2, sikdd} where we explored representations of documents which were valid over multiple languages.  The representations could be interpreted as multi-lingual topics, which were then used as proxies to compute cross-lingual similarities between documents. To learn the representations, we use Wikipedia as a training corpus. Significantly, we do not only consider the major or \emph{hub} languages such as English, German, French, etc. which have significant overlap in article coverage, but also smaller languages (in terms of number of Wikipedia articles) such as Slovenian and Hindi, which may have a negligible overlap in article coverage. We can then define a similarity between any two articles regardless of language,  which allows us to cluster the articles according to topic. The underlying assumption being that articles describing the same event are similar  and will therefore be put into the same cluster.

@@

\section{System Overview}

Event registry -> Monolingual events
Events -> features (semantic + CCA) -> linked events

@@

\subsection{Related work}

TODO

@@

We will fist introduce some notation regarding the training data, which we will use in the next subsections.

We will then present three approaches: a simple approach based on $k$-means clustering, a standard approach based on singular value decomposition and a new method that is applicable in more general
settings. All approaches are based on linear maps and we do not focus on alternatives, such as machine translation and probabilistic models. Applicability of machine translation relies
on the availability of training resources (strongly aligned translation corpora), which are not available for many world languages, but there exist comparable corpora. Probabilistic
methods \todo{provide reasons for not using them: e.g. difficulty with parameter estimation}.

For completeness we will also present the method of Canonical Correlation Analysis, since it is related to our proposed method.

\subsection{Notation}

@@

We now present the first approach.

\subsection{$k$-means}

@@

\subsection{Cross-Lingual Latent Semantic Indexing}

@@

The original sum of squared correlation is formulated as an unconstrained problem:
\begin{equation*}
  \begin{aligned}
    & \underset{w_i \in \RR^{k}}{\text{maximize}}
    & & \sum_{i < j}^m  \rho(w_i^T Y_i, w_j^T Y_j)^2.
\end{aligned}
\end{equation*}
We solve a similar problem by restricting $i=1$ and omit optimizing over non-hub language pairs.
Let $D_{i,i} \in \RR^{k \times k}$ denote the empirical covariance of $\mathcal{Y}_i$ and $D_{i,j}$ denote the empirical cross-covariance computed based on $\mathcal{Y}_i$ and $\mathcal{Y}_j$. We solve the following constrained (unit variance constraints) optimization problem:
\begin{equation}\label{squaredCorHubOriginal}
  \begin{aligned}
    & \underset{w_i \in \RR^{k}}{\text{maximize}}
    & & \sum_{i = 2}^m  \left(w_1^T D_{1,i} w_i \right)^2
    & \text{subject to}
    & & w_i^T D_{i,i} w_i = 1, \quad\forall i = 1,\ldots, m.
\end{aligned}
\end{equation}
The constraints $w_i^T D_{i,i} w_i$ can be simplified by using the Cholesky decomposition $D_{i,i} = K_i^T \cdot K_i$ and substitution: $y_i := K_i w_i$. By inverting the $K_i$ matrices and defining  $G_i := K_1^{-T} D_{1,i} K_i^{-1}$, the problem can be reformulated:
\begin{equation}\label{squaredCorHub}
  \begin{aligned}
    & \underset{y_i \in \RR^{k}}{\text{maximize}}
    & & \sum_{i = 2}^m  \left(y_1^T G_{i} y_i \right)^2
    & \text{subject to}
    & & y_i^T y_i = 1, \quad\forall i = 1,\ldots, m.
\end{aligned}
\end{equation}
A necessary condition for optimality is that the derivatives of the Lagrangian vanish. The Lagrangian of (\ref{squaredCorHub}) is expressed as:
$$  L(y_1, \ldots, y_m, \lambda_1, \ldots, \lambda_m) = \sum_{i = 2}^m  \left(y_1^T G_{i} y_i \right)^2 + \sum_{i=1}^m \lambda_i \left(y_i^T y_i - 1\right).$$
Stationarity conditions give us:
\begin{equation}\label{dLdx1}
 \frac{\partial}{\partial x_1} L = 0 \Rightarrow \sum_{i = 2}^m  \left(y_1^T G_{i} y_i \right) G_i y_i + \lambda_1 y_1,
\end{equation}
\begin{equation}\label{dLdxi}
\frac{\partial}{\partial x_i} L = 0 \Rightarrow \left(y_1^T G_{i} y_i \right) G_{i}^T y_1 + \lambda_i y_i = 0,~i > 1.
\end{equation}
Multiplying the equations (\ref{dLdxi}) with $y_i^T$ and applying the constraints, we can eliminate $\lambda_i$ which gives us:
\begin{equation}\label{eqy1yi}
G_{i}^T y_1 = \left(y_1^T G_{i} y_i \right) y_i,~i > 1.
\end{equation}
Plugging this into (\ref{dLdx1}), we obtain an eigenvalue problem:
$$\left( \sum_{i = 2}^m G_i G_{i}^T \right) y_1 + \lambda_1 y_1 = 0.$$
The eigenvectors of $\left( \sum_{i = 2}^m G_i G_{i}^T \right)$ solve the problem for the first language. The solutions for $y_i$ are obtained from (\ref{eqy1yi}): $y_i := \frac{G_{i}^T y_1}{\| G_{i}^T y_1 \|}$.
Note that the solution (\ref{squaredCorHubOriginal}) to can be recovered by: $w_i := K_i^{-1} y_i$. The linear transformation of the $w$ variables are thus expressed as:
$$ Y_1 := \text{eigenvectors of} \sum_{i = 2}^m G_i G_{i}^T, $$
$$ W_1 = K_1^{-1} Y_1 $$
$$ W_i = K_i^{-1} G_{i}^T Y_1 N,$$
where $N$ is a diagonal matrix that normalizes $G_{i}^T Y_1$, with $N(j,j) := \frac{1}{\|G(_{i} Y_1(:,j)\|}$.

@@

@@

\section{Evaluation}

We describe the datasets involved in the experiments and two evaluations: comparing the crosslingual approaches on an information retrieval task, and evaluation of cross-lingual event linking.

\subsection{Wikipedia Information Retrieval}

Mate retrieval results

\subsection{Experiments With Missing Alignment Data}

\subsection{Cross-Lingual Event Linking}

@@
